\section{Evaluation and Outlook}
\label{sec:eval}

\subsection{Review of the current offering}

\begin{itemize}
\item[(a)] We began the paper with the reconstruction of certain
mathematical objects, showing the technical feasibility of the
approach. 
\item[(b)] The more advanced example at the centre of the paper
  illustrates how this sort of reconstruction relates to mathematical
  practice.
\item[(c)] A future-oriented example exposes some technical
  challenges, while suggesting that blending could offer a novel
  approach to computer mathematics.
\end{itemize}

\subsection{Broader issues in evaluation}

In addition to motivating a further investigation of the role blending
can play in proofs, Galois theory, discussed above, is paradigmatic
for other reasons.  This discussion draws on the early 20th Century
writings of Albert Lautman on the philosophy of mathematics and the
subsequent interpretation of this work by Gilles Deleuze.  It uses
these ideas to propose an approach to embedding evaluation within the
system itself.

As Lautman writes in ``Ascent towards the Absolute'' \cite[p. 126]{lautman2011mathematics}:
\begin{quote}
What is characteristic of the movement of the theories that will be
considered here is the existence of an end conceived in advance as a
term of the ascent.
\end{quote}

This is reminiscent of our notion of coherence criteria that apply to
the blend.  To illustrate, let us briefly imagine how we would use
blending techniques to move from porcupine+lion to the perfected
\emph{porculione}.  Here, instead of field automorphisms that preserve
mathematical structure and fix certain designated elements, we would
look for mappings that preserve other properties that exist in the
underlying domain.  Porculiones would presumably have four feet, would
be mammals, and would be omnivores; they should also be viable living
creatures.

\cite[pp. 227--228]{deleuze1994difference} follows
Lautman in enthusiastically endorsing the Galoisian approach to
mathematics:
\begin{quote}
[T]he fact that an equation cannot be solved algebraically, for
example, is no longer discovered as a result of empirical research or
by trial and error, but as a result of the characteristics of the
groups and partial resolvents which constitute the synthesis of the
problem and its conditions (an equation is solveable only by algebraic
means -- in other words, by radicals, when the partial resolvents are
binomial equations and the indices of the groups are prime numbers).
The theory of problems is completely transformed and at last grounded,
since we are no longer in the classic master-pupil situation where the
pupil understands and follows a problem only to the extent that the
master already knows the solution and provides the necessary
adjunctions.  For, as Georges Verriest remarks, the group of an
equation does not characterise at a given moment what we know about
its roots, but the objectivity of what we do not know about them.
Conversely, this non-knowledge is no longer a negative or an
insufficiency but a rule or \emph{something to be learnt} which
corresponds to a fundamental dimension of the object.
\end{quote}

Although there is a commonality between blending and the
Galoisian approach insofar as progressive refinement carries us toward a
``perfected'' conclusion, Deleuze's enthusiasm about the pedagogical
situation would be significantly cooled here.  It would seem, in many
of our examples, that we only make progress ``to the extent that the
master already knows the solution and provides the necessary
adjunctions.''

However, this apparent infelicity may be less of a thick obstacle than
it would initially appear.  What seems to be most needed is a notion of a \emph{question} inside the system.  This would recover Lautman's basic
thrust: ``Scientific or not, every question has built in some
assumptions about the form of the answer'' \cite{larvor2011albert}.
In short, an experimental approach that allows the system to ask
and answer questions would embed key aspects evaluation in the system itself.

\subsection{Future work}

The idea of blending as steps in a proof would provide a useful
training ground for further development.  The primary problem is: If
blending is the realisation of ``combinatorial creativity'' how will
we avoid being swamped by the combinatorial explosion of possible
things to combine?
%
Here we can draw inspiration from two other recent projects that deal with
related concerns.
\cite{AbrSad14} develops a category theoretical approach to linguistic
anaphora that could be useful for the purpose of fitting different
mathematical components together in a sensible manner.
\cite{DBLP:journals/corr/GanesalingamG13} is a recently developed
system that uses a Logic of Computable Functions approach to prove
simple theorems, and that writes up its proofs in natural language.
%
Both systems are able to manage potential combinatorial explosion
in lingustic (and mathematical) content in a systematic manner.
%
A system that could selectively experiment with the rules it uses and
learn to associate different techniques with different types of
problems would be a useful synthesis and extension of these ideas.

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "mathsICCC"
%%% End: 
